Assessing the Global COVID19 Impact on Air Transport with Open Data

### 2021-06-18 paper.Rmd
good to loop

@@

1.2.1 Gather and prepare data
```{r}
#Reading and saving data
source("./R/list_apt_files.R")
file_names <- list_apt_files()
open_sky <- map_dfr(file_names, read_csv)
#write_csv(open_sky, "./data/open_sky.csv")
```

```{r}
# Take a look
glimpse(open_sky)
str(open_sky)
summary(open_sky)

# First exploration
fb1 <- open_sky %>% transmute(typecode = as.factor(typecode), origin = as.factor(origin), destination = as.factor(destination), firstseen, lastseen)
str(fb1)
summary(fb1)
# More than one million NA's

fb2 <- fb1 %>% drop_na()
str(fb2)
summary(fb2)
# Better...

#Visualizing something...
df1 <- fb2 %>% group_by(DATE = date(firstseen)) %>% summarize(count = n())
df1 %>% ggplot(aes(x = as.factor(year(DATE)))) +
  geom_boxplot(aes(y = count))
# This is the daily traffic of Aprils, distributed as a boxplot, FOR THE WHOLE WORLD! Needs to filter.

#Filtering Brazilian and European data to reduce the sample (and csv) size
#Preparing airports file

bra_10_apts <- c("SBBR", "SBGR", "SBSP", "SBKP", "SBRJ", "SBGL", "SBCF", "SBSV", "SBPA", "SBCT")
eur_apts <- c("EHAM","LFPG","EGLL","EDDF","EDDM","LEMD","LIRF","LEBL","EGKK","LSZH")
study_airports <- c(bra_10_apts, eur_apts)
study_airports

fb2 <- fb2 %>% filter(origin %in% bra_10_apts | destination %in% bra_10_apts) %>% mutate(REG = "BRA", .before = typecode)
fb2
summary(fb2)

#airports <- read_csv("data-raw/airports.csv")
#airports <- airports %>% transmute(ICAO = ident, CONTINENT = continent, COUNTRY = iso_country, REGION = iso_region, NAME #= name, CITY = municipality) 
#airports <- airports %>% mutate(CONTINENT = case_when(is.na(CONTINENT) ~ "NA", TRUE ~ CONTINENT))
#airports
#write_csv(airports, "./data/airports_info.csv")

@@

>>>>

```{r}
#If someone needs filters below, here we can control any selective sample
#Filtering Brazilian and European data to reduce the sample (and csv) size
#Currently, those filters are not in use in the code, they are here just in case.

bra_10_apts <- c("SBBR", "SBGR", "SBSP", "SBKP", "SBRJ", "SBGL", "SBCF", "SBSV", "SBPA", "SBCT")
eur_apts <- c("EHAM","LFPG","EGLL","EDDF","EDDM","LEMD","LIRF","LEBL","EGKK","LSZH")
study_airports <- c(bra_10_apts, eur_apts)
```

@@

1.2.1 Gather and prepare data
```{r}
#Reading raw data

source("./R/list_apt_files.R")

#Here I will assign only one month - "202105". If you want to include a full year, just assign year to "2021" or "2020". It works.
year <- "202105"
file_names <- list_apt_files(.year = year)
open_sky <- map_dfr(file_names, read_csv)

#Selecting relevant variables

fb <- open_sky %>% transmute(ADEP = origin, ADES = destination, TYPE = as.factor(typecode), DATE = date(day), CALL = callsign, ACFT_ID = aircraft_uid)

#Easily "dropping NA's# - this can be further sofisticated

fb1 <- fb %>% drop_na()
fb1
```

```{r}
#Joining countries and regions
# NOTE: the airports.csv file below is the downloaded file from www.ourairports.org.

airports <- read_csv("data-raw/airports.csv") %>% mutate(continent = case_when(is.na(continent) ~ "NA", TRUE ~ continent))
apt_countries <- airports %>% transmute(ICAO = ident, CTRY = iso_country)

#There are missing airports
apt_countries <- apt_countries %>% add_row(ICAO  = c("SPJC", "YBMC", "LSZM", "YSCH", "EPLB", "K3M3", "VV01", "SC28", "CWF2", "EHDB", "74XA", "HE13"), CTRY  = c("PE", "AU", "FR", "AU", "PL", "US", "VN", "US", "CA", "NL", "US", "EG"))

#If you need to write
#write_csv(apt_countries, "./data/apt_countries.csv")

#Joining to ADEP
fb2 <- left_join(fb1, apt_countries, by = c("ADEP" = "ICAO")) %>% mutate(ADEP_CTRY = CTRY, .keep = "unused", .after = ADEP)
#Joining to ADES
fb3 <- left_join(fb2, apt_countries, by = c("ADES" = "ICAO")) %>% mutate(ADES_CTRY = CTRY, .keep = "unused", .after = ADES)

#Check  NA's
colSums(is.na(fb3))
# Very few NA's - it's safe to drop and factor now

fb4 <- fb3 %>% drop_na() %>% mutate(ADEP_CTRY = as.factor(ADEP_CTRY), ADES_CTRY = as.factor(ADES_CTRY))

# Associate the regions
a <- airports %>% filter(continent == "EU") %>% select(iso_country) %>% unique()
eur_countries <- a$iso_country

#NOTE: Here we can adjust the European countries that comprises EU region by editing the "eur_countries" vector:
# Currently in european countries: "GB" "AD" "ES" "AL" "XK" "AT" "BA" "BE" "BG" "IS" "BY" "UA" "CH" "CZ" "SK" "DE" "RU" "DK" "HR" "EE" "FI" "GG" "JE" "IM" "NL" "IE" "FO" "LU" "NO" "PL" "PT" "SE" "LV" "LT" "FR" "GR" "HU" "IT" "LI" "SI" "MT" "MC" "RO" "TR" "MD" "MK" "GI" "RS" "ME" "SM" "GE" "VA"

base_dataset <- fb4 %>% mutate(ADEP_REG = as.factor(case_when(ADEP_CTRY == "US" ~ "US",
                                    ADEP_CTRY == "BR" ~ "BR",
                                    ADEP_CTRY %in% eur_countries ~ "EU",
                                    TRUE ~ "Other")), .after = ADEP_CTRY) %>%
                mutate(ADES_REG = as.factor(case_when(ADES_CTRY == "US" ~ "US",
                                    ADES_CTRY == "BR" ~ "BR",
                                    ADES_CTRY %in% eur_countries ~ "EU",
                                    TRUE ~ "Other")), .after = ADES_CTRY)
colSums(is.na(base_dataset))
# No NA's - yaaayy!!
glimpse(base_dataset)
summary(base_dataset)

@@

```{r}
#quick peek
base_dataset
#
#
#
#NOW IT'S TIME TO LOOP TO OTHER MONTHS AND BIND THEM ALL!
#
#
#

#quick first look
temp1 <- base_dataset %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n()) %>% mutate(ROUTE = paste(ADEP_REG, ADES_REG, sep = "-"), .keep = "unused")

temp1 %>% ggplot(aes(x = DATE)) +
  geom_line(aes(y = (FLIGHTS), color = ROUTE))
```
